Combining Static and Dynamic Analyses to Secure Programs against Integer Overflows

### 2012-09-02 doc/Manuscript/CGO/paper.tex
Added number of instructions necessary for instrumentation.

% What are integer overflows.
The most popular programming languages, including C, C++ and Java, limit the
size of primitive numeric types.
For instance, the \texttt{int} type, in Java, ranges from $-2^{31}-1$ to
$2^{31}$.
Consequently, there exist numbers that cannot be represented by these types.
In general, these programming languages resort to a {\em wrapping-arithmetics}
semantics~\cite{TODO} to perform integer operations.
If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
value wraps around, and $n \mbox{module} T_{max}$ wounds up represented
instead.
There are situations in which this semantics is acceptable~\cite{TODO}.
For instance, programers might rely on this behavior to implement hash functions
and random number generators.
On the other hand, there exist also situations in which this behavior might
lead a program to produce unexpected results.
As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
overflow -- a bug that resulted in a loss of more than U\$370 million
dollars~\cite{Dowson97}.
% TODO We need to explain better the wrapping-arithmetics, and we need to cite it
% properly. What could be a good (official) citation?
% And we need to cite

@@

% Our experimental results.
We use our range analysis to reduce the runtime overhead imposed by a dynamic
instrumentation library.
This instrumentation framework, also a contribution of this paper, has been
implemented in the LLVM compiler~\cite{Lattner04}.
We have used it to log overflows in a vast number of programs, and in this
paper we focus on SPEC CPU 2006.
Our implementation has been able to reproduce the results recently obtained by XX
{\em et al.}~\cite{TODO}.
Our range analysis algorithm has been able to remove XX\% of the overflow
checks created by the dynamic instrumentation framework.
All our experiments use an interprocedural version of range analysis.
We inline small functions to make our implementation more context sensitive.
Our experiments show that the speed and memory consumption of our range analysis
grows linearly with the program size.
By comparing the ranges that we determine statically with the results found by
a profiler, we see that more than 50\% of the bounds that we found are within a
factor of 50\% of the values that variables actually assume.

@@

We will use the program in Figure~\ref{fig:ex1}(a) to illustrate our range
analysis.
Figure~\ref{fig:ex1}(b) shows the same program in e-SSA form~\cite{Bodik00},
and Figure~\ref{fig:ex1}(c) outlines the constraints that we extract from this
program.
There is a clear correspondence between instructions and constraints.
Our analysis is sparse~\cite{Choi91}; thus, we associate one, and only one,
constraint with each integer variable defined in the program.
A possible solution to the range analysis problem, as obtained via the
techniques that we will introduce in Section~\ref{sub:algo}, is given in
Figure~\ref{fig:ex_eSSA_cgo}(d).

@@

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass;
thus, we work at the level of the compiler's intermediate
representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
This is in contrast to previous work, which either transforms the
source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
We work at the intermediate representation level to be able to couple our
library with static analyses, such as the algorithm that we described in
Section~\ref{sec:range}.
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after those instructions. 
The LLVM IR has five instructions that may lead to an overflow:
\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} and \textsc{Shl}
(left shift).
Figure~\ref{fig:instrumentation} shows the dynamic tests that we perform to
detect overflows.

>>>>

% What are integer overflows.
The most popular programming languages, including C, C++ and Java, limit the
size of primitive numeric types.
For instance, the \texttt{int} type, in Java, ranges from $-2^{31}-1$ to
$2^{31}$.
Consequently, there exist numbers that cannot be represented by these types.
In general, these programming languages resort to a {\em wrapping-arithmetics}
semantics~\cite{Warren02} to perform integer operations.
If a number $n$ is too large to fit into a primitive data type $T$, then $n$'s
value wraps around, and $n \mbox{module} T_{max}$ wounds up represented
instead.
There are situations in which this semantics is acceptable~\cite{Dietz12}.
For instance, programers might rely on this behavior to implement hash functions
and random number generators.
On the other hand, there exist also situations in which this behavior might
lead a program to produce unexpected results.
As an example, in 1996, the Ariane 5 rocket was lost due to an arithmetic
overflow -- a bug that resulted in a loss of more than U\$370 million
dollars~\cite{Dowson97}.
% TODO We need to explain better the wrapping-arithmetics, and we need to cite it
% properly. What could be a good (official) citation?
% And we need to cite

@@

% Our experimental results.
We use our range analysis to reduce the runtime overhead imposed by a dynamic
instrumentation library.
This instrumentation framework, also a contribution of this paper, has been
implemented in the LLVM compiler~\cite{Lattner04}.
We have used it to log overflows in a vast number of programs, and in this
paper we focus on SPEC CPU 2006.
Our implementation has been able to reproduce the results obtained by
previous work.
We have re-discovered the integer overflows recently observed by
Dietz {\em et al.}~\cite{Dietz12}, and the performance of our instrumentation
library, without the support of range analysis, matches the state-of-the-art
implementation of Brumley {\em et al.}
Our range analysis algorithm has been able to remove XX\% of the overflow
checks created by the dynamic instrumentation framework.
All our experiments use an interprocedural version of range analysis.
We inline small functions to make our implementation more context sensitive.
Our experiments show that the speed and memory consumption of our range
analysis grows linearly with the program size.
By comparing the ranges that we determine statically with the results found by
a profiler, we see that more than 50\% of the bounds that we found are within a
factor of 50\% of the values that variables actually assume.

@@

We will use the program in Figure~\ref{fig:ex_eSSA_cgo}(a) to illustrate our
range analysis.
Figure~\ref{fig:ex_eSSA_cgo}(b) shows the same program in e-SSA
form~\cite{Bodik00},
and Figure~\ref{fig:ex_eSSA_cgo}(c) outlines the constraints that we extract
from this program.
There is a clear correspondence between instructions and constraints.
Our analysis is sparse~\cite{Choi91}; thus, we associate one, and only one,
constraint with each integer variable defined in the program.
A possible solution to the range analysis problem, as obtained via the
techniques that we will introduce in Section~\ref{sub:prop}, is given in
Figure~\ref{fig:ex_eSSA_cgo}(d).

@@

%What is our instrumentation library
We have implemented our instrumentation library as a LLVM transformation pass;
thus, we work at the level of the compiler's intermediate
representation~\footnote{\texttt{http://llvm.org/docs/LangRef.html}}.
This is in contrast to previous work, which either transforms the
source code~\cite{Dietz12}, or the machine dependent code~\cite{Brumley07}.
We work at the intermediate representation level to be able to couple our
library with static analyses, such as the algorithm that we described in
Section~\ref{sec:range}.
Our instrumentation works by identifying the instructions that may 
lead to an overflow, and inserting assertions after those instructions. 
The LLVM IR has five instructions that may lead to an overflow:
\textsc{Add}, \textsc{Sub}, \textsc{Mul},  \textsc{Trunc} (also bit-casts) and
\textsc{Shl} (left shift).
Figure~\ref{fig:instrumentation} shows the dynamic tests that we perform to
detect overflows.
Instrumenting \textsc{Add} or \textsc{Sub} requires 12 LLVM instructions.
Left shifts need 8 instructions, \textsc{Mul} requires 7 and
\textsc{Trunc} requires only 3.
